ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}

### 2021-09-09 manuscript/main.tex
update manuscript, scripts and figures to final submission (last reviewer comment addressed, editorial requests addressed)

\usepackage{xr-hyper}

@@

\usepackage[left]{lineno}
\linenumbers

@@

\title{ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
\author[1,2,3,4]{David Lähnemann}
\author[4,5]{Johannes Köster}
\author[3]{Ute Fischer}
\author[3]{Arndt Borkhardt}
\author[1,2,+,*]{Alice C. McHardy}
\author[5,6,+,*]{Alexander Schönhuth}
\affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
\affil[2]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[3]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[4]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
\affil[5]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
\affil[6]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
\affil[+]{shared last authors}
\affil[*]{To whom correspondence should be addressed:
 \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de},
 \href{as@cwi.nl}{as@cwi.nl}.}

@@

\section*{\Large Abstract}
{\sffamily
Obtaining accurate mutational profiles from single cell DNA is essential for the analysis of genomic cell-to-cell heterogeneity at the finest level of resolution.
However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which---when applied to single cells---generate significant numbers of false positives and false negatives.
Only dedicated models accounting for amplification bias and errors will be able to provide more accurate calls.

We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
ProSolo models a single cell jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered sites, when downstream tools cannot handle missing data.
ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to quantify, such as allele dropout.

ProSolo's model is implemented into a flexible framework, encouraging extensions. The source code and usage instructions are available at: \url{https://github.com/prosolo/prosolo}
}

\section{Introduction}

@@

We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
Importantly, our model enables a scalable---because computationally efficient---implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.

\section{Results}

\subsection{Single Cell Sequencing Model}

We describe a novel probabilistic model that addresses the genotyping of diploid single cells whose DNA has been subject to whole genome multiple displacement amplification (MDA)\cite{dean_comprehensive_2002}.
In the following, we introduce the central innovations of our model and demonstrate its advantages in comparison to existing approaches.
More details and a detailed derivation of all model elements can be found in the Supplement, including a summary of the core model.

Our model addresses the two major issues of MDA: (i) the differential amplification of the two alleles present in a diploid cell ("amplification bias" in the following); (ii) MDA induced errors ("amplification errors" in the following) which are copy errors introduced by the $\Phi29$ polymerase used in MDA. To address amplification bias, we leverage a mechanistically motivated, empirically derived model of differential amplification of alleles.
To assess amplification errors, we evaluate single cell samples together with a bulk sample from which the single cell is supposed to stem.
Regarding the latter, we argue that a bulk sample should be added to single cell sequencing experiments wherever possible: it samples from the same cell population without requiring amplification, and is therefore unaffected by amplification bias and errors and thus makes a particularly useful background sample to address the statistical uncertainties and biases induced by MDA.
At the same time, one of the major features of the core model and its implementation is that it can easily be adapted to flexibly deal with other sampling setups, so it could be extended to further scenarios.
For related work on flexible bulk sequencing sample composition, see Köster et al.\cite{koster_varlociraptor_2020}.

\begin{figure}[!tpb]%figure1
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_hom-ref-model_cov-20.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_het-model_cov-20.pdf}
    \end{minipage}
  \end{minipage}
  \vspace{2ex}
  \hspace{1ex}
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/prosolo_alt-calling_overview.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.49\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Event_space.pdf}
    \end{minipage}
  \end{minipage}
  \caption{
   \textbf{A,B} Exemplary alternative allele read count distributions for sites covered by 20 reads, as derived by Lodato et al.\cite{lodato_somatic_2015}
   Homozygous reference sites in \textbf{(A)} are assumed to follow a beta-binomial distribution;
   sites heterozygous for the alternative allele in \textbf{(B)} are assumed to follow the linear combination of two symmetrical beta-binomial distributions (dotted and dashed lines).
   \textbf{(C)} Toy example of calling the same genomic site in two single cells from the same population that differ in their true underlying allele frequencies for alternative allele C (blue, $\theta_s = 0$ vs. $\theta_s = 0.5$).
   Alternative nucleotide T (orange) is an amplification error.
   Empirical distributions in A and B account for the  amplification bias, and likelihoods for the alternative allele candidates from the bulk reduce the likelihoods of amplification errors, thereby correctly identifying both the error and the original true mutation.
   This is formalised with the model in D.
   \textbf{(D)} Definition of single cell events based on ProSolo's likelihood density estimates for the spectrum of true underlying alternative allele frequencies in the single cell ($\tilde{\theta}_s$) and the bulk ($\tilde{\theta}_b$).
   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
   The model further assumes that the bulk sample has sufficient coverage to capture somatic variants.\small\newline
   ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
  \label{fig:prosolo_alt-calling}
\end{figure}

the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}C).
We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele, to calculate likelihoods for each of these possible true allele frequencies.

To do this, we follow the considerations of Lodato et al.\cite{lodato_somatic_2015} (see Figure S5 and the section ``Modeling MDA-derived alternative read counts'' of the respective supplement for the details), who fitted well-studied probability distributions to empirical distributions they obtained for $\mathbf{P}(\rho_s\mid\theta_s)$ of $\theta_s\in\{0,0.5\}$.
For sites that are homozygous for the reference allele, amplification bias cannot initially happen.
However, once an amplification error creates an alternative allele, this can be amplified to large frequencies due to amplification bias.
Lodato et al.\cite{lodato_somatic_2015} thus consider the empirical distribution $\mathbf{P}(\rho_s\mid\theta_s=0)$ to follow a beta-binomial.
Effectively, this means that the probability of a non-zero alternative read count  ($\rho_s>0$, which is any alternative read count above $0$ in Figure~\ref{fig:prosolo_alt-calling}A) will be non-zero ($\mathbf{P}(\rho_s\mid\theta_s=0)>0$), merely because of sequencing and amplification errors.
In contrast, at heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}B), the distribution is dominated by amplification bias.
Thus, for $\mathbf{P}(\rho_s\mid\theta_s=0.5)$ they found a mixture of two beta-binomial distributions to appropriately fit the empirically observed distributions (Figure~\ref{fig:prosolo_alt-calling}B).
We further motivate the choice of the beta-binomial distribution mechanistically by an analogy to its generative urn model named after Pólya\cite{eggenberger_uber_1923}:
Take an urn with two white and two black balls, where each ball represents one strand of each (double-stranded) allele at a heterozygous site.
In the Pólya urn model, drawing a ball leads to replacement of that ball with two balls of the same color, analogous to a strand copy by the $\Phi29$ polymerase (additional discussion in the Supplement, Section~\ref{sec:polya-urn}).
Finally, the necessity for a mixture of beta-binomials (Figure~\ref{fig:prosolo_alt-calling}B) becomes evident when contrasting it with the binomial distribution observed at heterozygous sites in bulk experiments.
Namely, the bulk distribution for reads supporting the alternative allele would narrowly peak at a count of $k=\frac{l}{2}$, which is not the case for the corresponding $k=10$ in Figure~\ref{fig:prosolo_alt-calling}B.
Instead, the mixture of beta-binomials peaks at the extreme read counts of $k=0$ and $k=l$, highlighting that the dropout of the alternative or reference allele is quite likely to occur.
Most likely, this mixture arises from differences in the accessibility of different pieces of the genomic DNA for amplification\cite{picher_trueprime_2016} (Supplement, Section~\ref{sec:polya-urn}).
Finally, for homozygous alternative sites (i.e. $\mathbf{P}(\rho_s\mid\theta_s=1)$), we rely on the symmetry of the cases $\theta_s=0$ and $\theta_s=1$.
In summary, we obtain the following equations for $\mathbf{P}(\rho_s\mid\theta_s)$:

Here, $\cB\cB$ represents the beta-binomial probability mass function where $k\in\{0,...,l\}$ with $l$ being the total read coverage of the site that is considered. All of parameters $\alpha, \alpha_1,\alpha_2$, $\beta$, and $w$ scale linearly in $l$.
Allowing to vary distributions through these parameters allows amplification bias to depend on the total coverage of a site, and thereby to vary locally.
Symmetry of distributions for the homozygous sites ($\theta_s = 0$ in Figure~\ref{fig:prosolo_alt-calling}A, and $\theta_s = 1$) is established by swapping the shape parameters $\alpha(l)$ and $\beta(l)$.
The distribution for heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}B) corresponds to a mixture of two beta-binomials with shape parameters $\alpha_1(l)=\beta_2(l)$ and $\alpha_2(l)=\beta_2(l)$, where equality of $\alpha(l)$ and $\beta(l)$ yields symmetry of the distributions relative to $k = \frac{l}{2}$ (the expected value for alternative read counts at heterozygous sites).
Slopes and intersects for the scaling of all parameter values across different choices of $l$ are given in Table~\ref{tab:Lodato}.

\paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
A bulk sequencing sample of the same cell type from the same organism is a much larger sampling of that cell population than the sequencing of dozens of single cells.
Unless a single cell is the only one to harbour a particular mutation, a deep enough sequencing of the bulk sample from which the single cell was drawn should contain reads from cells that share the particular mutation with the single cell.
Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
This establishes a formal, statistical argument why one should consider bulk experiments in single cell sequencing whenever possible: the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell, in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
At the same time, the absence of a single cell candidate mutation in a bulk sample (with sufficient sequencing depth) increases the probability of an amplification error.
As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.

\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across all possible allele frequencies $\tilde{\theta}_s$ and $\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:

For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel; for discussion on the validity of this assumption, see Supplementary Section \ref{sec:event-space-def}).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
For example, an error-free homozygous alternative site is defined by allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
And the dropout of an alternative allele is defined across allele frequency likelihood density estimates $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.

We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}D):

Assuming a flat prior across the possible underlying allele frequencies for both the bulk and the single cell, we can compute likelihoods for all those single cell events (e.g.~Equation~\ref{eq:ado-to-ref-likelihood-data}).
The sum of the likelihoods of all these (mutually exclusive) events yields the marginal probability (Equation \ref{eq:marginal-prob}).
Using the marginal probability, we can calculate the posterior probability for any of these events.
For example, the posterior probability of event $\boldsymbol{E}_{\text{hom alt}}$ (Figure~\ref{fig:prosolo_alt-calling}D) can be calculated with:

With read probabilities calculated with Equation \ref{eq:read_probability}, we can calculate the posterior probability of the $\boldsymbol{E}_{\text{hom alt}}$ event (for an analogous and more detailed derivation for $\boldsymbol{E}_{\text{ADO to ref}}$, see Supplement, Section~\ref{sec:sc-model}).
To obtain posterior probabilities for compound events, we sum up the posterior probabilities of the events it comprises, effectively joining up their respective allele frequency range combinations into compound range combinations.
For the site-specific single cell probability for the presence of an alternative allele in the single cell, we thus get the compound event (blue events in Figure~\ref{fig:prosolo_alt-calling}D; Table~\ref{tab:Events}):

To genotype, we calculate the posterior probability of all three possible single-cell genotypes and choose the genotype with the maximum posterior probability. 
In Figure~\ref{fig:prosolo_alt-calling}D, events are colored purple when implying the homozygous reference genotype, light blue when implying the heterozygous genotype and dark blue when implying the homozygous alternative genotype (Table \ref{tab:Events}).
Similarly, to calculate the posterior probability of an allele dropout at a particular site, we sum up the posterior probabilities of the two {\ttfamily ADO} events.

\subsection{Benchmarking}

\subsubsection{Datasets and Ground Truths}

\begin{figure}[!tpb]
  \begin{minipage}{.3\linewidth}
    \includegraphics[height=38ex]{figs/Dong2017_data_overview.pdf} \\
  \end{minipage}
  \hspace{6ex}
  \begin{minipage}{.5\linewidth}
    \includegraphics[height=38ex]{figs/Laehnemann2017_data_overview.pdf} \\
  \end{minipage}\\
  \begin{minipage}{.5\linewidth}
    \includegraphics[height=38ex]{figs/Wang2014_data_overview.pdf} \\
  \end{minipage}
  \hspace{6ex}
  \begin{minipage}{.3\linewidth}
    \includegraphics[height=38ex]{figs/data_overview_color_legend.pdf} \\
  \end{minipage}
  \caption{
   Benchmark Datasets.
   For single cells, DNA was multiple displacement amplified (MDA).
   \textbf{A} Whole genome sequencing (WGS) dataset, generated from a clonal population started with an individual cell\cite{dong_accurate_2017}, and expanded further downstream to generate different bulk samples (C1, C2, C3, IL-1C).
   \textbf{B} Newly generated whole exome sequencing (WXS) dataset of blood cells.
   Ground truth genotypes of patient CCS1 were determined from sequencing data of family members (boxes male; circles female).
   Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
   \textbf{C} Whole exome sequencing (WXS) dataset from a triple negative breast cancer (TNBC) patient\cite{wang_clonal_2014}.
   For tumor cells (a1 to a8 and h1 to h8), the cell population (pop) sequencing of the normal sample was taken as the ground truth, including known clonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
   For normal cells (n1 to n16), the cell population (pop) sequencing of the tumor sample was taken as the ground truth, after removing known clonal and subclonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
   }
  \label{fig:datasets}
\end{figure}

\paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from SRA project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
After an initial mini-expansion, a single cell was selected as the founder for the secondary IL expansion into 20-30 cells.
From this, two cells were extracted (IL-11 and IL-12) and sequenced following MDA.
The remaining kindred cells from that clone were used as a bulk sequencing sample without amplification (IL-1C).
IL-1C serves as the ground truth, as these cells are only very few cell divisions away from IL-11 and IL-12, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated using GATK HaplotypeCaller to call variant sites and bcftools mpileup to identify homozygous reference sites (with read coverage above 25 but no alternative allele present).
IL-1C was only used as a ground truth and not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3), generated from cells after the first mini-expansion, were merged into a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
Unlike other callers (all of which finished in less than 5 days), SCIPhI took five weeks to finish on this dataset in sensitive mode and 7.5 weeks in default mode.\\

\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
\begin{sloppypar}
For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}B, dataset available as \mbox{EGAD00001005929} in EGA study \mbox{EGAS00001004123}).
Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
Using a panel of 16 primer pairs covering different genes across chromosomes for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.
\end{sloppypar}

To generate the ground truth of this dataset, we could leverage previously sequenced bulk whole exome data from the same person, their parents and three siblings \cite{hoell_constitutional_2014} (Figure~\ref{fig:datasets}B).
To create ground truth germline alternative allele calls, we ran three pedigree-aware variant callers (BEAGLE4.0\cite{browning_improving_2013}, polymutt\cite{li_likelihood-based_2012} and FamSeq\cite{peng_rare_2013,peng_famseq:_2014}) and created a consensus by including only calls where all callers agree at a site and where a maximum of one caller not calling the site was allowed (Figure~\ref{fig:granulocytes-ground-truth}).

\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC)\cite{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available (Figure~\ref{fig:datasets}C, dataset available via the sequence read archive, project id \mbox{PRJNA168068} or accession \mbox{SRA053195}).
For each, 16 single cell nuclei and a bulk sample were whole exome sequenced.
Further, a number of candidate sites for clonal and subclonal tumor mutations was confirmed by targeted duplex deep sequencing in a separate experiment\cite{wang_clonal_2014}.
We analyzed the tumor and normal cells separately, ensuring that normal cells were called with the normal bulk background sample and tumor cells with the tumor bulk background sample.
As the ground truth for the tumor cells, we then used the normal bulk sample augmented with the clonal tumor mutations confirmed by targeted duplex sequencing.
As the ground truth for the normal cells, we used the tumor bulk sample, removing the confirmed clonal and subclonal tumor mutations.
This experimental setup aims for fairness across all competitors. 
Variant calling for the ground truths was performed as for cell line data above.

\subsubsection{ProSolo achieves consistently high alternative allele calling accuracy}

The most precise single cell variant callers to date, SCcaller and SCIPhI, only call the presence vs. the absence of an alternative allele (i.e. the heterozygous and the homozygous alternative genotypes called jointly, the joint probability of the blue event areas in Figure~\ref{fig:prosolo_alt-calling}D).
We thus focused on this for the main benchmarking.

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
 \end{minipage}\newline
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=40ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}

 \caption{
  Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
  All panels are strong zoom-ins, focusing on (different) areas of interest.
  Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B. \newline \footnotesize
  Threshold parameters (not comparable across tools):
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
\label{fig:alt-calling_prec-rec}
\end{figure}

>>>>

\usepackage{xr-hyper}
\usepackage{hyperref}

@@

%\usepackage[left]{lineno}
%\linenumbers

@@

\title{ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
\author[1,2,3,4,5]{David Lähnemann}
\author[5,6]{Johannes Köster}
\author[4]{Ute Fischer}
\author[4]{Arndt Borkhardt}
\author[1,2,3,*]{Alice C. McHardy}
\author[6,7,*]{Alexander Schönhuth}
\affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
\affil[2]{Braunschweig Integrated Centre of Systems Biology (BRICS), Technische Universität Braunschweig, Braunschweig, Germany.}
\affil[3]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[4]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[5]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
\affil[6]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
\affil[7]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
\affil[*]{These authors jointly supervised this work:
 \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de},
 \href{as@cwi.nl}{as@cwi.nl}.}

@@

\section*{\Large Abstract}
{\sffamily
Accurate single cell mutational profiles can reveal genomic cell-to-cell heterogeneity.
However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
The resulting data violates assumptions of variant callers developed for bulk sequencing.
Thus, only dedicated models accounting for amplification bias and errors can provide accurate calls.

We present ProSolo for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
ProSolo probabilistically models a single cell jointly with a bulk sequencing sample and integrates all relevant MDA biases in a site-specific and scalable---because computationally efficient---manner.
This achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools and supports imputation of insufficiently covered genotypes, when downstream tools cannot handle missing data.
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly.

ProSolo is implemented in an extendable framework, with code and usage at: \url{https://github.com/prosolo/prosolo}
}

\section*{Introduction}

@@

In this work we describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
Importantly, our model enables a scalable---because computationally efficient---implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.
It achieves a higher variant calling accuracy compared to state-of-the-art tools.

\section*{Results}

\subsection*{Comprehensive and Flexible Single Cell Sequencing Model}

We describe a probabilistic model that addresses the genotyping of diploid single cells whose DNA has been subject to whole genome multiple displacement amplification (MDA)\cite{dean_comprehensive_2002}.
Here, we name the central innovations of our model and demonstrate its advantages in comparison to existing approaches.
A more detailed description of the innovations is available in the Methods section.

Briefly, our model addresses the two major issues of MDA: (i) the differential amplification of the two alleles present in a diploid cell ("amplification bias" in the following); (ii) MDA induced errors ("amplification errors" in the following) which are copy errors introduced by the $\Phi29$ polymerase used in MDA.
To address amplification bias, we leverage a mechanistically motivated, empirically derived model of differential amplification of alleles.
To assess amplification errors, we evaluate single cell samples together with a bulk sample from which the single cell is supposed to stem.
Regarding the latter, we argue that a bulk sample should be added to single cell sequencing experiments wherever possible: it samples from the same cell population without requiring amplification, and is therefore unaffected by amplification bias and errors and thus makes a particularly useful background sample to address the statistical uncertainties and biases induced by MDA.
At the same time, one of the major features of the core model and its implementation is that it can easily be adapted to flexibly deal with other sampling setups, so it could be extended to further scenarios.

\subsection*{Consistently High Alternative Allele Calling Accuracy}

The most precise single cell variant callers to date, SCcaller and SCIPhI, only call the presence vs. the absence of an alternative allele (i.e. the heterozygous and the homozygous alternative genotypes called jointly, the joint probability of the blue event areas in Figure~\ref{fig:prosolo_alt-calling}d).
We thus focused on this for the main benchmarking.

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
 \end{minipage}\newline
 \begin{minipage}{.95\linewidth}
  \includegraphics[height=45ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}

 \caption{
  Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
  All panels are strong zoom-ins, focusing on (different) areas of interest.
  Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
  \textbf{a} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
  \textbf{b} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
  \textbf{c} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel b. \newline \footnotesize
  Threshold parameters (not comparable across tools):
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
\label{fig:alt-calling_prec-rec}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}a, Methods section) seems much less challenging than the other dataset, as all methods achieved very high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec}a and \ref{fig:alt-calling_prec-rec_global}a), at recall rates of .45 and higher. 
In comparison to all other tools, ProSolo shows striking increases in recall.
For example, an increase of nearly 10\% for a precision above .99, where its maximum recall is .766, compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
The only exception with a recall of .0001 is SCAN-SNV (at a precision of .992).
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset (Figure \ref{fig:datasets}b, Methods section).
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}b and \ref{fig:alt-calling_prec-rec_global}b).
In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
